One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-07-29 papers/1_dmatpol/paper-r3.tex
Revision

In this work, the electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities
interacting with electric fields. The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The electric field\hyp{}induced interaction energies,
dipole moments and quadrupole moments
are quantitatively described by the present theory in weak, moderate and even very strong electric fields (0.003 -- 0.07 a.u.).
It is believed that the proposed model could open
new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems
at arbitrary level of theory.
%at post\hyp{}Hartree\hyp{}Fock levels of theory.
\end{abstract}

@@

and consider electric field magnitudes in a range from very weak (~0.003 a.u.) to strong (0.07 a.u.).

@@

in weak, moderate and strong electric fields.}
In this Table, `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields,
respectively. $R$-squared coefficients are shown in parentheses.
Ranks of the density matrix susceptibility models are denoted as $R_{\rm max}$
where $R_{\rm max}$ is the maximum power in the expansion
from Eq.~\eqref{e:final-model.General}.

@@

\multicolumn{15}{c}{\textbf{Non-Uniform Electric Field}} \\ 

@@

S & 78.07 & (40.53) & - & (-) & - & (-) && 43.33 & (81.68) & 14.40 & (97.98) && 64.33 & (59.62) \\ \hline

@@

\end{table*}

@@

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{data/dmatpol/water/figure1/fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the density matrix polarization susceptibility tensors of water molecule
in uniform electric field.} The notation $(R=R_{\max})$ denotes the maximum rank of the electric field
in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0.0 to 0.07 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
\end{figure}
All the tested models that are linear with respect to the electric field,
i.e., Eq.~\eqref{e:final-model.HF} as well as Eqs.~\eqref{e:dD-Taylor} and \eqref{e:final-model.General} 
truncated at linear susceptibilities, ${\bf B}^{(1)}$,
incorrectly reproduce the polarization energy.
Interestingly, the predicted values are in noticeable correlation with the exact estimates with a slope of $\sim$2.
Incorporation of the second\hyp{}order term
obtained either by numerical differentiation from Eq.~\eqref{e:B20} (red filled circles) 
or by using the generalized DMS model (red open circles)
dramatically reduces the errors
leading to quantitative accuracy in all tested field magnitudes 
with the best RMSE of 0.173 kcal/mol ($R^2=99.1\%$) for the generalized DMS model.
On the other hand, the
induced dipole and quadrupole moments are reproduced reasonably well already by
using the first\hyp{}order DMS, ${\bf B}^{(10)}$, with $R^2$ coefficients being 
greater than 97\% in case of dipole moments and grater than 80\% in case of quadrupole moments.
This observation can be understood by the fact that the induced dipole
and quadrupole moment is approximately proportional to the external electric field, whereas the polarization
energy exhibits quadratic dependence (c.f. Eq.~\eqref{e:dE-dpol}), hence the need of including at least
quadratic terms in the expansion from Eq.~\eqref{e:dD-Taylor}.

@@

The most sophisticated model studied in this work, the generalized quadratic DMS model, yields excellent predictions 
of the polarization energies and induced multipole moments in all the electric field strengths tested.
The errors in predicting polarization energy are reduced by an order of magnitude when compared with
the linear DMS models. Improvement of accuracy can also be observed in the case of induced multipole moment predictions
but the effect is less significant (improvement roughly by a factor of 2).
What is particularly interesting is that in weak and moderate electric fields
the generalized quadratic DMS model performs better than the distributed dipole\hyp{}dipole
polarizability model, whereas in strong electric fields the situation is reversed
(RMSE of 0.402, $R^2=97.2\%$ and 0.332 kcal/mol, $R^2=98.2\%$ for the DMS and polarizability model, respectively).
Nevertheless, the quadratic DMS model yields still very high $R^2$ of 97.2\% in strong electric fields for polarization
energy, and even higher $R^2$ for induced multipole moments.
Recently, it was found by Fried et~al. by performing molecular dynamics simulations with a classical 
polarizable force field, that the average electric field experienced by water molecule in liquid phase
in room temperature is around 0.014 a.u. ($\sim$70 MV/cm).\cite{Fried.Wang.Boxer.Ren.Pande.JPCB.2013}
This value corresponds to our moderate electric field set of samples (denoted as `M' in Tables~\ref{t:fields}
and \ref{t:results}) for which the
accuracy of the generalized quadratic DMS model is excellent
Water molecules can experience
stronger electric fields in certain geometries of hydrogen bonded clusters, 
even up to 0.055 a.u. according to study by Reischl et~al.\cite{Reischl.Kofinger.Dellago.MolPhys.2009} 
This electric field magnitude falls into our strong electric field regime in which the generalized quadratic
DMS model still provides accurate predictions (see also error distributions 
shown in Figure~\ref{f:fig-2}B, \ref{f:fig-2}E and \ref{f:fig-2}H). 
It is emphasized here that the current model can be easily extended by taking into account
third\hyp{}order terms. 
%This however falls out of scope of this contribution and it will be studied
%elsewhere. 
Therefore, it is believed that
the generalized DMS approach 
has an important advantage over the multipole\hyp{}based approach, i.e., it provides detailed
tensor description of the spatial polarization of the electron density distribution, that could be
combined with density\hyp{}based methods of interaction energy calculations.\cite{Mandado.Hermida-Ramon.JCTC.2011}

>>>>

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities.
The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially uniform electric field, as well as
in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The electric field\hyp{}induced polarization energies,
dipole moments and quadrupole moments are quantitatively described by the present theory 
in weak, moderate and even very strong electric fields (0.003 -- 0.07 a.u.)
when quadratic terms with respect to the electric field are accounted for.
It is believed that the proposed model could open
new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
\end{abstract}

@@

and consider electric field magnitudes in a range from very weak (~0.003 a.u.) to very strong (0.07 a.u.).

@@

in uniform and non\hyp{}uniform electric fields.$^{a,b}$}

@@

\multicolumn{15}{c}{\textbf{Non-Uniform Electric Field}$^{c}$} \\ 

@@

S & 78.07 & (40.53) & - & (-) & - & (-) && 43.33 & (81.68) & 14.40 & (97.98) && 64.33 & (59.62) \\ 

@@

@@

\footnotesize{\flushleft{
$^a$ $R$-squared coefficients are shown in parentheses. `n.d.' stands for `not determined'. \\
$^b$ Ranks of the density matrix susceptibility models are denoted as $R_{\rm max}$ where $R_{\rm max}$ is the maximum power in the expansion from Eq.~\eqref{e:final-model.General}.\\
$^c$ `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields 
that are generated by sets of point charges, respectively (c.f. Table~\ref{t:fields}).\\
}}

\end{table*}
%
\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{data/dmatpol/water/figure1/fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the density matrix polarization susceptibility tensors of water molecule
in uniform electric field.} The notation $(R=R_{\max})$ denotes the maximum rank of the electric field
in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0.0 to 0.07 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
} 
\end{figure}
%
All the tested models that are linear with respect to the electric field,
i.e., Eq.~\eqref{e:final-model.HF} as well as Eqs.~\eqref{e:dD-Taylor} and \eqref{e:final-model.General} 
truncated at linear susceptibilities, ${\bf B}^{(1)}$,
incorrectly reproduce the polarization energy.
Interestingly, the predicted values are in noticeable correlation with the exact estimates with a slope of $\sim$2.
Incorporation of the second\hyp{}order term
obtained either by numerical differentiation from Eq.~\eqref{e:B20} (red filled circles) 
or by using the generalized DMS model (red open circles)
dramatically reduces the errors
leading to quantitative accuracy in all tested field magnitudes 
with the best RMSE of 0.173 kcal/mol ($R^2=99.1\%$) for the generalized DMS model.
On the other hand, the
induced dipole and quadrupole moments are reproduced reasonably well already by
using the first\hyp{}order DMS, ${\bf B}^{(10)}$, with $R^2$ coefficients being 
greater than 97\% in case of dipole moments and grater than 80\% in case of quadrupole moments.
This observation can be understood by the fact that the induced dipole
and quadrupole moment is approximately proportional to the external electric field, whereas the polarization
energy exhibits quadratic dependence (c.f. Eq.~\eqref{e:dE-dpol}), hence the need of including at least
quadratic terms in the expansion from Eq.~\eqref{e:dD-Taylor}.

@@

The most sophisticated model studied in this work, the generalized quadratic DMS model, yields excellent predictions 
of the polarization energies and induced multipole moments in all the electric field strengths tested.
The errors in predicting polarization energy are reduced by an order of magnitude when compared with
the linear DMS models. Improvement of accuracy can also be observed in the case of induced multipole moment predictions
but the effect is less significant (improvement roughly by a factor of 2).
What is particularly interesting is that in weak and moderate electric fields
the generalized quadratic DMS model performs better than the distributed dipole\hyp{}dipole
polarizability model, whereas in strong electric fields the situation is reversed
(RMSE of 0.402, $R^2=97.2\%$ and 0.332 kcal/mol, $R^2=98.2\%$ for the DMS and polarizability model, respectively).
Nevertheless, the quadratic DMS model yields still very high $R^2$ of 97.2\% in strong electric fields for polarization
energy, and even higher $R^2$ for induced multipole moments.
Recently, it was found by Fried et~al. by performing molecular dynamics simulations with a classical 
polarizable force field, that the average electric field experienced by water molecule in liquid phase
in room temperature is around 0.014 a.u. ($\sim$70 MV/cm).\cite{Fried.Wang.Boxer.Ren.Pande.JPCB.2013}
This value corresponds to our moderate electric field set of samples (denoted as `M' in Tables~\ref{t:results}
and \ref{t:fields}) for which the
accuracy of the generalized quadratic DMS model is excellent.
Water molecules can experience
stronger electric fields in certain geometries of hydrogen bonded clusters, 
even up to 0.055 a.u. according to study by Reischl et~al.\cite{Reischl.Kofinger.Dellago.MolPhys.2009} 
This electric field magnitude falls into our strong electric field regime in which the generalized quadratic
DMS model still provides accurate predictions (see also error distributions 
shown in Figure~\ref{f:fig-2}B, \ref{f:fig-2}E and \ref{f:fig-2}H). 
It is emphasized here that the current model can be easily extended by taking into account
third\hyp{}order terms. 
%This however falls out of scope of this contribution and it will be studied
%elsewhere. 
Therefore, it is believed that
the generalized DMS approach 
has an important advantage over the multipole\hyp{}based approach, i.e., it provides detailed
tensor description of the spatial polarization of the electron density distribution, that could be
combined with density\hyp{}based methods of interaction energy calculations.\cite{Mandado.Hermida-Ramon.JCTC.2011}
